Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-11 appendix.tex
change notation in plotting code; gradient argument for lemma 3

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the divergence term
\[
\shannonDivergence_{\tau^*,t^*}(\tau^*,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau^*,t) \mid_{t=t^*} = 0.
\]

Assume $\emptyset$ is the most likely ancestral state for all site splits.
In this case, the partial likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.

@@

We focus on $x_1$, similar arguments applying to the other variables.
The partial likelihood
\[
\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t)
\]
as a function of $x_1$, using Table~\ref{tab:sitepatprob}, is
\[
\tilde{\ell}(x_1) \propto \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1).
\]
By some calculus, the partial derivative with respect to $x_1$ is
\[
\tilde{\ell}_{x_1}(x_1) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
\]
which is clearly negative for all $x_1 \ge x^*$.
%TODO: make more explicit/clear why this means \hat{x}_1 \in [0,x^*]
This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in [0, x^*], \ \hat{y}_1 \in [0, y^*], \ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

@@

\begin{proof}
%TODO: make this more explicit/clear
Let
\[
a_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
\]
with similar definitions for the remaining site splits.
The objective function proportional to $w$ is
\begin{align*}
\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + a_{\siteSplit_j}w)
\end{align*}
and has derivative with respect to $w$ as
\begin{align*}
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{1 + a_{\siteSplit_j}w}.
\end{align*}
Note the following:
\begin{itemize}
\item $a_{\emptyset} = -a_{13}$, $p_{\emptyset} \ge p_{13}$;
\item $a_{1} = -a_{3}$, $p_{1} = p_{3}$;
\item $a_{2} = -a_{123}$, $p_{2} = p_{123}$;
\item $a_{12} = -a_{23}$, $p_{12} = p_{23}$;
\end{itemize}
so
\begin{align*}
\ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}a_{\emptyset}}{1 + a_{\emptyset}w} - \frac{p_{13}a_{\emptyset}}{1 - a_{\emptyset}w} + \frac{p_{2}a_{2}}{1 + a_{2}w} - \frac{p_{2}a_{2}}{1 - a_{2}w} \\
         &\qquad \frac{p_{1}a_{1}}{1 + a_{1}w} - \frac{p_{1}a_{1}}{1 - a_{1}w} + \frac{p_{12}a_{12}}{1 + a_{12}w} - \frac{p_{12}a_{12}}{1 - a_{12}w} \\
\end{align*}
Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{y}_1$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1$ and $x_2 \ge x_1$, we have
\[
0 \le a_{\emptyset} \le \frac{2x^*}{1+(x^*)^2},
\]
\[
0 \le a_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*},
\]
\[
0 \le a_{1} \le \frac{x^*}{1-(x^*)^2}
\]
\[
0 \le a_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*},
\]
so that $\ell'(w)$ has lower bound
\[
\ell'_{L}(w) = \frac{1}{1+w} - \frac{p_{13}}{c_1 - w} - \frac{p_{2}}{c_1c_2 - w} - \frac{p_{1}}{c_3 - w} - \frac{p_{12}}{c_2c_3 - w}
\]
with
\[
c_1 = \frac{1+(x^*)^2}{2x^*},
\]
\[
c_2 = 1-y^*,
\]
\[
c_3 = \frac{1-(x^*)^2}{x^*}.
\]
Clearly $c_1 > 1$.
We see
\[
\ell'_{L}(0) = 1 - \frac{p_{13}}{c_1} - \frac{p_{2}}{c_1c_2} - \frac{p_{1}}{c_3} - \frac{p_{12}}{c_2c_3} > 0
\]
provided $c_1c_2, c_3, c_2c_3 > 1$
Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$.
Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1)$, including at the boundary $w=1$, implying $\hat{w} \equiv 1$.
So we want $x^*$ and $y^*$ where $c_1c_2, c_3, c_2c_3 > 1$ and $\ell_L'(1) > 0$.
These conditions are straightforward, the one on the derivative being
\[
\ell'_{L}(1) = \frac{1}{2} - \frac{p_{13}}{c_1-1} - \frac{p_{2}}{c_1c_2-1} - \frac{p_{1}}{c_3-1} - \frac{p_{12}}{c_2c_3-1} > 0.
\]
The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state is given in Figure~\ref{fig:incons-analytic}.
\end{proof}

>>>>

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the divergence term
\[
\shannonDivergence_{\tau^*,t^*}(\tau^*,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the divergence, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to Table~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
The unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau^*,t) \mid_{t=t^*} = 0.
\]
We now show the gradient of the divergence term is a decreasing function on $(0,1)$ in each variable.
We focus on $x_1$.
The divergence proportional to $x_1$ can be written as
\begin{align*}
\shannonDivergence(x_1) \propto \sum_j p_{\siteSplit_j} \log(c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
\end{align*}
where $c_{\siteSplit_j}$ and $a_{\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example,
\[
c_{\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\shannonDivergence'(x_1) := \frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}
\end{align*}
and focus on a single term
\[
f_j(x_1) = p_{\siteSplit_j}a_{\siteSplit_j}\frac{1}{c_{\siteSplit_j} + a_{\siteSplit_j}x_1}.
\]
Regardless of $a_{\siteSplit_j} \lessgtr 0$, this is a decreasing function of $x_1$ provided $c_{\siteSplit_j} + a_{\siteSplit_j}x_1 > 0$, but we see
\[
p_{\siteSplit_j} := \frac{1}{8} (c_{\siteSplit_j} + a_{\siteSplit_j}x_1)
\]
and $p_{\siteSplit_j} > 0$ so long as $x^*\in(0,1)$ and $y^*\in(0,1)$.

We now focus on the partial likelihood term.
Assume $\emptyset$ is the most likely ancestral state for all site splits.
In this case, the partial likelihood has no ambiguity, and has terms given in Table~\ref{tab:likelihoods-restricted}.

@@

The partial likelihood
\[
\tilde{\ell}_{\tau^*,t^*}(\tau^*, t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau^*, t)
\]
as a function of $x_1$, using Table~\ref{tab:sitepatprob}, is
\[
\tilde{\ell}(x_1) \propto \left(\frac{1}{2}+\frac{1}{2}x^*(y^*)^2\right)\log(1+x_1) + \left(\frac{1}{2}-\frac{1}{2}x^*(y^*)^2\right)\log(1-x_1).
\]
By some calculus, the partial derivative with respect to $x_1$ is
\[
\tilde{\ell}_{x_1}(x_1) = \frac{-x_1+x^*(y^*)^2}{1-x_1^2}
\]
which is clearly negative for all $x_1 \ge x^*$.
%TODO: make more explicit/clear why this means \hat{x}_1 \in [0,x^*]
This same process applied to the remaining variables shows the gradient of the full likelihood is negative for $x_1,x_2 \ge x^*$ and $y_1 \ge y^*$ and positive for $y_2,w \ge y^*$, yielding the constraints
\[
\hat{x}_1,\hat{x}_2 \in [0, x^*], \ \hat{y}_1 \in [0, y^*], \ \hat{y}_2,\hat{w} \in [y^*, 1].
\]
\end{proof}

@@

\begin{proof}
%TODO: make this more explicit/clear
Similar to the previous proof, let
\[
b_{\emptyset} = \frac{(x_1+x_2)(y_1+y_2)}{(1+x_1x_2)(1+y_1y_2)}
\]
with similar definitions for the remaining site splits.
The objective function proportional to $w$ is
\begin{align*}
\ell(w) \propto \log(1+w) + \sum_j p_{\siteSplit_j} \log(1 + b_{\siteSplit_j}w)
\end{align*}
and has derivative with respect to $w$ as
\begin{align*}
\ell'(w) := \frac{d}{dw} \ell(w) = \frac{1}{1+w} + \sum_j p_{\siteSplit_j}b_{\siteSplit_j}\frac{1}{1 + b_{\siteSplit_j}w}.
\end{align*}
Note the following:
\begin{itemize}
\item $b_{\emptyset} = -b_{13}$, $p_{\emptyset} \ge p_{13}$;
\item $b_{1} = -b_{3}$, $p_{1} = p_{3}$;
\item $b_{2} = -b_{123}$, $p_{2} = p_{123}$;
\item $b_{12} = -b_{23}$, $p_{12} = p_{23}$;
\end{itemize}
so
\begin{align*}
\ell'(w) &= \frac{1}{1+w} + \frac{p_{\emptyset}b_{\emptyset}}{1 + b_{\emptyset}w} - \frac{p_{13}b_{\emptyset}}{1 - b_{\emptyset}w} + \frac{p_{2}b_{2}}{1 + b_{2}w} - \frac{p_{2}b_{2}}{1 - b_{2}w} \\
         &\qquad \frac{p_{1}b_{1}}{1 + b_{1}w} - \frac{p_{1}b_{1}}{1 - b_{1}w} + \frac{p_{12}b_{12}}{1 + b_{12}w} - \frac{p_{12}b_{12}}{1 - b_{12}w} \\
\end{align*}
Using the bounds on $\hat{x}_1,\hat{y}_1,\hat{y}_1$, and $\hat{y}_2$ from Lemma~\ref{lemma:bounds} and our assumption of $y_2 \ge y_1$ and $x_2 \ge x_1$, we have
\[
0 \le b_{\emptyset} \le \frac{2x^*}{1+(x^*)^2} := b_{\emptyset}^U,
\]
\[
0 \le b_{2} \le \frac{2x^*}{1+(x^*)^2}\cdot\frac{1}{1-y^*} := b_{2}^U,
\]
\[
0 \le b_{1} \le \frac{x^*}{1-(x^*)^2} := b_{1}^U
\]
\[
0 \le b_{12} \le \frac{x^*}{1-(x^*)^2}\cdot\frac{1}{1-y^*} := b_{12}^U,
\]
so that $\ell'(w)$ has lower bound
\[
\ell'_{L}(w) = \frac{1}{1+w} - \frac{p_{13}}{1/b_{\emptyset}^U - w} - \frac{p_{2}}{1/b_{2}^U - w} - \frac{p_{1}}{1/b_{1}^U - w} - \frac{p_{12}}{1/b_{12}^U - w}
\]
Clearly $b_{\emptyset}^U < 1$.
We see
\[
\ell'_{L}(0) = 1 - p_{13}b_{\emptyset}^U - p_{2}b_{2}^U - p_{1}b_{1}^U p_{12}b_{12}^U > 0
\]
provided $b_{2}^U, b_{1}^U, b_{12}^U < 1$.
Note that $\ell'_{L}(w)$ is strictly decreasing in $w$ on the interval $(0,1)$, as shown in the previous proof.
Thus, if $\ell'_{L}(1) > 0$ then $\ell(w)$ is increasing on the entire interval $(0,1)$, including at the boundary $w=1$, implying $\hat{w} \equiv 1$.
So we want $x^*$ and $y^*$ where $b_{2}^U, b_{1}^U, b_{12}^U < 1$ and $\ell_L'(1) > 0$.
These conditions are straightforward, the one on the derivative being
\[
\ell'_{L}(1) = \frac{1}{2} - \frac{p_{13}b_{\emptyset}^U}{1-b_{\emptyset}^U} - \frac{p_{2}b_{2}^U}{1-b_{2}^U} - \frac{p_{1}b_{1}^U}{1-b_{1}^U} - \frac{p_{12}b_{12}^U}{1-b_{12}^U} > 0.
\]
The set of $x^*,y^*$ satisfying these conditions as well as those for $\emptyset$ being the most likely ancestral state is given in Figure~\ref{fig:incons-analytic}.
\end{proof}
